---
title: "R Midterm take home portion"
author: "Harsh Tandon W1580393"
output:
  html_document:
    df_print: paged
---

```{r}
dir()
setwd("D:/1st Qtr Study Material/R/Midterm")  #set working directory
#import all used libraries
library(dplyr)
library(ggplot2)
library(jsonlite)
```

# Part 1
### 1. Load titanic.csv dataset
```{r}
data = read.csv("D:/1st Qtr Study Material/R/Midterm/titanic.csv")
```

### 2. Get rid of columns "Name", "Ticket", "PassengerId", "Cabin"
```{r}
colToRemove = c(which(colnames(data)=="Name"),
                which(colnames(data)=="Ticket"),
                which(colnames(data)=="PassengerId"),
                which(colnames(data)=="Cabin"))

data = data[-colToRemove]
```

### 3. Replace the ‘Sex column with a binary column for whether the passenger is 'Male'
```{r}
levels(data$Sex) = c(0,1)
levels(data$Sex) 

colIndexOfSex = which(colnames(data)=="Sex") 
names(data)[colIndexOfSex] = "isMale"
data
```

### 4. Show how many NA’s are in each column 
```{r}
colSums(is.na(data))
```

### 5. Which Column has the most NA’s? How many does it have? 
We see from Q4 that "Age" has most number of NA's. It has 177 NA values 

### 6. Create a bar chart to show NA counts by column
```{r}
NAcount = (colSums(is.na(data)))
barplot(NAcount, xlab = "Column Name", ylab = "Count of NA", ylim = c(0,200), main = "NA counts by columns")
```

### 7. Plot the distribution of ages for each gender. 
#### Is the average (mean) age of Males higher? 
From the density plot, we see that Average age of Males is higher than Females. 
```{r}
mu <- data %>% 
  group_by(isMale) %>%
  summarise(grp_mean = mean(Age, na.rm = TRUE))
mu
ggplot(data = data, aes(x = Age)) + 
  geom_density(aes(fill = isMale), alpha = 0.4) + 
  geom_vline(aes(xintercept = grp_mean, color = isMale),
             data = mu, linetype = "dashed",size = 1)
```

#### Is the median higher? 
We see that Median value of Age for Males is higher than Females.
```{r}
median <- data %>% 
  group_by(isMale) %>%
  summarise(grp_median = median(Age, na.rm = TRUE))
median
ggplot(data = data, aes(x = Age)) + 
  geom_density(aes(fill = isMale), alpha = 0.4) + 
  geom_vline(aes(xintercept = grp_median, color = isMale),
             data = median, linetype = "dashed",size = 1)
```

### 8. Which port of origin (embarked column) had the highest average fare? Show/plot your results.
```{r}
avg_Fare <- data %>% 
  group_by(Embarked) %>%
  summarise(grp_mean = mean(Fare, na.rm = TRUE))
avg_Fare

barplot(avg_Fare$grp_mean, names.arg = avg_Fare$Embarked, main = "Average Fare by Port", ylab = "Average Fare", xlab = "Port")
```

### 9. Is there a statistically significant difference in average fares between the genders? 
No, the difference between the average fares of Males and Females is statistically significant. 
Average for Males: 25.52
Average for Females: 44.48
The difference is 18.96!
```{r}
avg_Fare <- data %>% 
  group_by(isMale) %>%
  summarise(grp_mean = mean(Fare, na.rm = TRUE))
avg_Fare

barplot(avg_Fare$grp_mean, names.arg = avg_Fare$isMale, main = "Average Fare by Gender", ylab = "Average Fare", xlab = "isMale", ylim = c(0,50))

```

# Part 2
### 1. Load the bitly data from data.gov as a data frame. 
```{r}
data2 = stream_in(url("https://raw.githubusercontent.com/denisvrdoljak/MSIS2506_Fall2019/master/TakeHomeMidterm/bitly_usagov_example.txt"))
```

### 2. How many records are there? 
```{r}
print(nrow(data2))
```

### 3. In the City column (‘cy’), how many are NA? 
```{r}
sum(is.na(data2$cy))
```

### 4. How many countries are present in this data? 
#### Which country (‘c’ column) has the most records? How many does it have? 
From the results we see that, US has the most records. It has 2305 records!
```{r}
data2$c = as.factor(data2$c)
levels(data2$c)
summary(data2$c)
```

### 5. How many records are from Russia? 
There are 13 records from Russia.
```{r}
count(data2,  data2$c == "RU")
```

### 6. How many records do NOT list USA?
There are 614 records that do not list USA.
```{r}
count(data2, data2$c != "US")
```

### 7. How many records have 'America' in the timezone, but Country not in US? 
There are 143 records having "America" in timezone, but country not equal to US.
```{r}
hasAmerica = dplyr::filter(data2, grepl("America", tz))
count(hasAmerica, c != "US")
```

### 8. How many records point to cia.gov? 
```{r}
cia = dplyr::filter(data2, grepl("cia.gov", u))
length(cia)
```

### 9. Fill empty records with “UNKNOWN” and NA’s with “MISSING”. Create a new csv. 
```{r}
sum(is.na(data2))
data2[is.na(data2)] = "MISSING"
data2[data2 == ""] = "UNKNOWN"
sum(is.na(data2))
data3 = apply(data2,2,as.character)

write.table(data3, file = "NewCsv.csv", sep = ",", col.names = TRUE, row.names = FALSE)
```

### 10.  Show the top 10 timezones (‘tz’ column) in a bar chart
```{r}
data2$tz = as.factor(data2$tz)
timezones = summary(data2$tz)
timezones = sort(timezones, decreasing = TRUE)
top10 = timezones[1:10]

par(mar=c(10,4,4,4))
barplot(top10, las = 2, main = "Top 10 Timezones", ylab = "Count", col = 1:6)
legend("topright", cex = 0.75, ncol = 2, legend = names(top10), fill = 1:6)
title(xlab="Timezones" ,mgp=c(9,1,0))
```
